Follow-up Samples

### 2016-12-06 submission/manuscript_outline_20161024.Rmd
manuscript method model update and mcnemar initial

***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
A total of ```r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))``` sample was omitted from the original ```r length(rownames(good_metaf))``` sample set since it was missing a complete set of follow up data.  This left a total of ```r sum(initial_cm$ref_yes)``` samples for test predictions.  After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4].  This decrease was significant for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)``` [Figure 4a] and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4b] alone.  This also held specifically for those with screen relative neoplasias (SRN) (P-value = ```r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)```).  A total of ```r initial_cm["pred_yes", "ref_yes"]``` or ```r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)```% of all samples were correctly predicted to have an adenoma or carcinoma.  Although there was a decrease in positive probability only ```r followup_cm["pred_no", "ref_no"]``` of the total ```r sum(initial_cm$ref_yes)``` individuals were classified as adenoma or carcinoma free on follow up (successful classification of ```r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)```%). There was no significant difference between the predictions and actual diagnosis as measured by the Mcnemar test (P-value = ```r confusion_summary["McnemarPValue", "initial"]```).  However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = ```r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)```).  Although there were discordent results the respective sensitivity for the initial group was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% and for follow up was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```%, respectively.

@@

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the ```r length(rownames(good_metaf))``` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v```r paste(packageVersion("caret"))```).  The method of choosing the best model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  This was repeated 100 times with the best model after all the repeats chosen to be used on the ```r length(rownames(good_metaf))``` samples with follow up.  The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as having or not having a lesion.

***Selection of Important OTUs***
In order to assess which variables were most central to all the models the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) was recorded and those varialbes that were present more than 50% of the time were considered important.

A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples and specifically lesion.  This new model OTUs were then compared to the overall lesion model and common OTUs were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.

>>>>

***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
A total of ```r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))``` sample was omitted from the original ```r length(rownames(good_metaf))``` sample set since it was missing a complete set of follow up data.  This left a total of ```r sum(initial_cm$ref_yes)``` samples for test predictions.  After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4].  This decrease was significant for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)``` [Figure 4a] and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4b] alone.  This also held specifically for those with screen relative neoplasias (SRN) (P-value = ```r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)```).  A total of ```r initial_cm["pred_yes", "ref_yes"]``` or ```r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)```% of all samples were correctly predicted to have an adenoma or carcinoma.  Although there was a decrease in positive probability only ```r followup_cm["pred_no", "ref_no"]``` of the total ```r sum(initial_cm$ref_yes)``` individuals were classified as adenoma or carcinoma free on follow up (successful classification of ```r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)```%). There was no significant difference between the predictions and actual diagnosis for the initial samples in the 67-sample cohort test set.  However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = ```r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)```).  Although there were discordent results the respective sensitivity for the initial group was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% and for follow up was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```%, respectively.

@@

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the ```r length(rownames(good_metaf))``` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v```r paste(packageVersion("caret"))```).  Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  This was repeated 100 times with the best model after all the repeats chosen to be used on the ```r length(rownames(good_metaf))``` samples with follow up.  Once the ideal mtry was found the entire ```r length(rownames(metaI))``` sample set was used to create the final Random Forest model.  The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.

***Selection of Important OTUs***
In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important.

A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.  This new model OTUs were then compared to the overall lesion model and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.

### 2016-12-06 submission/manuscript_outline_20161024.tex
manuscript method model update and mcnemar initial

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this study, we
assessed whether a model built with bacterial microbiome data could
accurately predict adenoma or carcinoma and adjust positive probability
for lesion (adenoma or carcinoma) after surgical treatment. This model
was tested on a \texttt{66} person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model chosen had a 10-fold cross validated AUC
of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
up samples our Random Forest model significantly decreased the positive
probability of lesion compared to the initial samples for both adenoma
(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
\texttt{5.96e-08}). In the test set, the prediction of lesion for the
initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
and \texttt{60.6}\% had a lesion. Our model suggests that treatment does
significantly reduce the probability of having a colonic lesion. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
Decrease in Positive Probability Prediction}} A total of \texttt{1}
sample was omitted from the original \texttt{67} sample set since it was
missing a complete set of follow up data. This left a total of
\texttt{66} samples for test predictions. After multiple comparison
correction there was a significant overall decrease in positive
probability of a carcinoma and adenoma (P-value = \texttt{8.27e-11})
{[}Figure 4{]}. This decrease was significant for both adenoma (P-value
= \texttt{8.48e-07} {[}Figure 4a{]} and carcinoma (P-value =
\texttt{5.96e-08}) {[}Figure 4b{]} alone. This also held specifically
for those with screen relative neoplasias (SRN) (P-value =
\texttt{1.45e-04}). A total of \texttt{65} or \texttt{98.5}\% of all
samples were correctly predicted to have an adenoma or carcinoma.
Although there was a decrease in positive probability only \texttt{26}
of the total \texttt{66} individuals were classified as adenoma or
carcinoma free on follow up (successful classification of
\texttt{40.9}\%). There was no significant difference between the
predictions and actual diagnosis as measured by the Mcnemar test
(P-value = \texttt{1}). However, the predictions were significantly
discordent with the diagnosis for the follow up samples (P-value =
\texttt{1.17e-09}). Although there were discordent results the
respective sensitivity for the initial group was \texttt{98.5}\% and for
follow up was \texttt{100}\%, respectively.

@@

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). We chose the middle training model, based on AUC, from 100
80/20 (train/test) splits. It's 10-fold cross validated AUC was similar
to it's test set AUC which was not the case for both the best and worse
training model {[}Figure 3{]}. Using this model we predicted the
probability of a lesion in the inital and follow up samples {[}Figure
4{]}. There was a signficant decrease in positive probability regardless
of whether the sample as a carcinoma or adenoma. The overall sensitivity
for lesion detection in the intial samples was \texttt{98.5} and for
follow ups was \texttt{100}. Although there was a decrease in overall
probability of an adenoma or carcinoma only \texttt{26} were below the
0.5 threshold out of the total \texttt{65} individuals who were
diagnosed as not having a carcinoma on follow up.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the \texttt{67} individuals with
follow up samples. The model included data on FIT, the bacterial
microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was
caucasion or not, history of cancer, and family history of cancer.
Non-binary data was checked for near zero variance and auto correlation.
Data columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v\texttt{6.0.73}). The method of
choosing the best model involved taking the \texttt{490} samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
twenty times through 10-fold cross validation to optimzie the model's
Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
This resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. This was repeated 100
times with the best model after all the repeats chosen to be used on the
\texttt{67} samples with follow up. The default cutoff of a probability
of 0.5 was used as the threshold to classify individuals as having or
not having a lesion.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models the number of times a
variable was present in the top 10\% of mean decrease in accuracy (MDA)
was recorded and those varialbes that were present more than 50\% of the
time were considered important.

A second model using the same types of conditions was used to identify
OTUs that could predict initial and follow up samples and specifically
lesion. This new model OTUs were then compared to the overall lesion
model and common OTUs were selected to test for changes between initial
and follow up using a paired wilcoxson rank sum test.

>>>>

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this study, we
assessed whether a model built with bacterial microbiome data could
accurately predict adenoma or carcinoma and adjust positive probability
for lesion (adenoma or carcinoma) after surgical treatment. This model
was tested on a \texttt{66} person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model chosen had a 10-fold cross validated AUC
of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
up samples our Random Forest model significantly decreased the positive
probability of lesion compared to the initial samples for both adenoma
(P-value = \texttt{7.95e-08}) and carcinoma (P-value =
\texttt{7.95e-08}). In the test set, the prediction of lesion for the
initial samples had a sensitivity of \texttt{100}\% (\texttt{66/66})
while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
predicted that \texttt{36.4}\% of the 67-person cohort had normal colons
and \texttt{63.6}\% had a lesion. Our model suggests that treatment does
significantly reduce the probability of having a colonic lesion. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
Decrease in Positive Probability Prediction}} A total of \texttt{1}
sample was omitted from the original \texttt{67} sample set since it was
missing a complete set of follow up data. This left a total of
\texttt{66} samples for test predictions. After multiple comparison
correction there was a significant overall decrease in positive
probability of a carcinoma and adenoma (P-value = \texttt{1.14e-11})
{[}Figure 4{]}. This decrease was significant for both adenoma (P-value
= \texttt{7.95e-08} {[}Figure 4a{]} and carcinoma (P-value =
\texttt{7.95e-08}) {[}Figure 4b{]} alone. This also held specifically
for those with screen relative neoplasias (SRN) (P-value =
\texttt{5.35e-04}). A total of \texttt{66} or \texttt{100}\% of all
samples were correctly predicted to have an adenoma or carcinoma.
Although there was a decrease in positive probability only \texttt{24}
of the total \texttt{66} individuals were classified as adenoma or
carcinoma free on follow up (successful classification of
\texttt{37.9}\%). There was no significant difference between the
predictions and actual diagnosis for the initial samples in the
67-sample cohort test set. However, the predictions were significantly
discordent with the diagnosis for the follow up samples (P-value =
\texttt{4.19e-10}). Although there were discordent results the
respective sensitivity for the initial group was \texttt{100}\% and for
follow up was \texttt{100}\%, respectively.

@@

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). We chose the middle training model, based on AUC, from 100
80/20 (train/test) splits. It's 10-fold cross validated AUC was similar
to it's test set AUC which was not the case for both the best and worse
training model {[}Figure 3{]}. Using this model we predicted the
probability of a lesion in the inital and follow up samples {[}Figure
4{]}. There was a signficant decrease in positive probability regardless
of whether the sample as a carcinoma or adenoma. The overall sensitivity
for lesion detection in the intial samples was \texttt{100} and for
follow ups was \texttt{100}. Although there was a decrease in overall
probability of an adenoma or carcinoma only \texttt{24} were below the
0.5 threshold out of the total \texttt{65} individuals who were
diagnosed as not having a carcinoma on follow up.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the \texttt{67} individuals with
follow up samples. The model included data on FIT, the bacterial
microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was
caucasion or not, history of cancer, and family history of cancer.
Non-binary data was checked for near zero variance and auto correlation.
Data columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v\texttt{6.0.73}). Optimiation of
the mtry hyperparameter as well as data on the best and worse
performance of the model involved taking the \texttt{490} samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
twenty times through 10-fold cross validation to optimzie the model's
Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
This resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. This was repeated 100
times with the best model after all the repeats chosen to be used on the
\texttt{67} samples with follow up. Once the ideal mtry was found the
entire \texttt{490} sample set was used to create the final Random
Forest model. The default cutoff of a probability of 0.5 was used as the
threshold to classify individuals as positive or negative of lesion.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important.

A second model using the same types of conditions was used to identify
OTUs that could predict initial and follow up samples. This new model
OTUs were then compared to the overall lesion model and the resulting
common OTUs to both models were selected to test for changes between
initial and follow up using a paired wilcoxson rank sum test.
